---
title: "Assignment 3: Data Exploration"
author: "Chunyi Xu"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}
#Noting down codes to install packages
#install.packages("tidyverse"); 
#install.packages("lubridate");
#install.packages("here")

#Loading necessary packages (tidyverse, lubridate, here)
library (tidyverse)
library (lubridate)
library (here)

#Checking working directory is the project folder
here ()

#Uploading the ECOTOX neonicotinoid dataset 
Neonics <- read.csv(
  file = here("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv"),
  stringsAsFactors = TRUE)   

#Uploading the Niwot Ridge NEON dataset for litter and woody debris
Litter <- read.csv(
  file = here("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv"),
  stringsAsFactors = TRUE)   


```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer:The ecotoxicology of neonicotinoids on insects helps us better understand their ecological impacts. Although neonicotinoid insecticides were introduced because their relatively lower toxicity than other older insecticides, recent research has indicated that their environmental impact may also be devastating. Neonicotinoid can have harmful consequences for benefical insects, pollinator, and aquatic invertebrates.
Source: https://www.xerces.org/pesticides/understanding-neonicotinoids#:~:text=While%20they%20were%20initially%20introduced,beneficial%20insects%2C%20and%20aquatic%20invertebrates.

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer:Litter and woody debris are essential components of forest and stream ecosystems. They are involved in the nutrient cycling and carbon budgets process, serving as a source of energy for aquatic ecosystems and providing habitat for aquatic and terrestrial organisms. They also influence the ecosystem's structure and roughness and could potentially change sediment transport and water flows. 
Source: https://research.fs.usda.gov/treesearch/20001



4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1.Litter and woody debris is sampled at terrestrial NEON sites that contain woody vegetation >2m tall.
 2.The sampling plot's centers should be 50m from buildings and other non-NEON infrastructure. 
 3. The placement of litter traps within the plots may be either targeted or randomized, depending on the vegetation. For example, in sites with more than 50% aerial cover of woody vegetation >2m in height, the trap placement is randomized. In sites with less than 50% cover of woody vegetation and distributed vegetation, the placement is targeted. 

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
#Checking the dimensions of the dataset Neonics. It has 4623 rows and 30 columns. 
dim(Neonics)

```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r}
# Creating a new dataframe "Effect_Summary" for the summary of the "Effect" column
Effect_Summary <- summary (Neonics$Effect)
# Viewing the new dataframe "Effect_Summary"
print (Effect_Summary)
# Sorting the new dataframe "Effect_Summary" in order of magnitude
sort (Effect_Summary)

```

> Answer:The top six effects that are studied are Population, Mortality, Behavior, Feeding behavior, Reproduction, and Development. These effects are studied because they have a more significant impact on the biodiversity of an ecosystem than other impacts. For example, population is essential to maintaining habitat diversity, sustaining the balance of the food chain, and controlling the carbon cycle. However, studies have found that the use of neonicotinoids leads to increased population extinction rates. Therefore, mortality is worth investigating, too, as it is closely linked to the population of a species. Animal behavior, including feeding behavior, can potentially change ecosystem dynamics, so this topic becomes specifically of interest, too. Animal behavior is critical to delivering ecosystem services, which determine the material benefits people receive, such as food and fuel, and the regulating benefits, such as climate regulation and flood control. 
Source: https://www.noble.org/regenerative-agriculture/wildlife/top-5-considerations-for-increasing-wildlife-diversity/
https://environment.yale.edu/news/article/protecting-wildlife-populations-can-enhance-natural-capture-capture 
https://www.sciencedirect.com/science/article/abs/pii/S0169534722002762 
https://aithor.com/essay-examples/an-analysis-of-the-impact-of-feeding-habits-on-animal-behavior-and-ecosystems#3-the-ecological-consequences-of-feeding-habits 

7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r}
# Creating a new dataframe "Effect_Species.Common.Name" for the summary of the "Effect" column
Effect_Species.Common.Name <- summary (Neonics$Species.Common.Name)
# Viewing the new dataframe "Effect_Species.Common.Name"
print (Effect_Species.Common.Name)
# Sorting the new dataframe "Effect_Species.Common.Name" in order of occurrence
sort (Effect_Species.Common.Name)

```

> Answer:The six most commonly studied species in the dataset are Honey Bee, Parasitic Wasp, Buff Tailed Bumblebee, Carniolan Honey Bee, Bumble Bee, and Italian Honeybee. These species are all bees. Bees are the primary pollinators in the ecosystem. They play important roles in maintaining and improving biodiversity by transporting flower pollen and fertilizing plants and have an impact on other insects, animals, and plants. Therefore, they are vital in sustaining the global food supply and a healthy ecosystem. However, a significant amount of studies indicate that neonics usage is a leading cause of massive bee die-offs around the world.
Source: https://greenly.earth/en-us/blog/ecology-news/why-are-bees-so-important-for-people-and-the-environment
https://www.nrdc.org/stories/neonicotinoids-101-effects-humans-and-bees

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r}
#Figuring out the class of "Conc.1..Author", and the class is "factor"
class (Neonics$Conc.1..Author)
#Viewing the dataframe
#View (Neonics)
#View(Neonics$Conc.1..Author)
```

> Answer: By viewing the entire dataframe Neonics, the entries in the column "Conc.1..Author" includes both values and symbols "/". When we uploaded the dataset, we read strings in as factors. Therefore, the column "Conc.1..Author" is classified as factor. Meanwhile, by viewing the column "Conc.1..Author" separately, the result also shows that the value for "Conc.1..Author" are non-numeric but factors with 1006 levels <0.0004 <0.025 <0.088 <0.5 <1.5 <10/ <2.5/ <4.00 <5.00 <8.79 ... NR/. 

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}
#Generating a plot for the number of studies conducted by publication year by using the dataset Neonics
ggplot(Neonics) +
geom_freqpoly(aes(x = Publication.Year), bins = 50)

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}
# Reproducing the same graph but adding a color aesthetic according to the variable Test.Location
ggplot(Neonics) +
geom_freqpoly(aes(x = Publication.Year, color = Test.Location), bins = 50) +
theme(legend.position = "left")

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer:The most common test location is the Lab. The second common test location is Field Natural. They differ over time. For example, from around 1993 to 2000, and from around 2007 to 2010, Field Nature was a more common test location than Lab. However, from around 2003 to 2005, and from around 2012 to 2020, Lab has become a more common test location than Field Natural. From 2000 to around 2003, both locations were equally common. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r}
# Creating a bar graph of Endpoint counts and adjusting the X-axis labels
ggplot(Neonics)+
geom_bar(aes(x = Endpoint))+
theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

> Answer: The two most common end points are NOEL and LOEL. LOEL is for the terrestrial database usage. It means lowest-observable-effect-level: lowest does (concentration) producing effects that were significantly different (as reported by authors) from responses of controls (LOEAL/LOEC). NOEL is for terrestrial database usage too. It means no-observable-effect-level: highest does (concentration) producing effects not significantly different from responses of controls according to author's reported statistical test (NOEAL/NOEC).

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}
# Determining the class of collectDate, and the result shows that it is not a date. It is "factor" instead. 
class (Litter$collectDate)

#Changing collectDate to a date 
Litter$collectDate <- as.Date(Litter$collectDate, format = "%y-%m-%d")

#Confirming the new class of collectDate is date
class (Litter$collectDate)

# Determining which dates litter was sampled in August 2018
unique (Litter$collectDate)


```

13. Using the `unique` function, determine how many different plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
#Using the unique function to determine how many different plots were sampled at Niwot Ridge (NIWO), and the result is 12
unique (Litter$plotID)

# Comparing the result given by the unique function and the summary function
summary (Litter$plotID)

```

> Answer: There are 12 different plots were sampled at the Niwot Ridge, since there 12 levels in the plotID column. While the information obtained from 'unique' only shows the unique values that one factor can take, the information obtained from 'summary' gives the detailed counts of each unique value. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}


```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}


```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer:

What type(s) of litter tend to have the highest biomass at these sites?

> Answer:
