---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Chunyi Xu"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
#Loading necessary packages (tidyverse, lubridate, here)
library (tidyverse)
library (lubridate)
library (agricolae)
library (ggplot2)
library(corrplot)
library (dplyr)
library (here)

#Checking working directory is the project folder
here()

#Importing the NTL-LTER raw data file
NTL_LTER <- read.csv(
  file = here("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"),
  stringsAsFactors = TRUE)   

#Setting date columns to date objects
NTL_LTER$sampledate <- as.Date(NTL_LTER$sampledate, format = "%m/%d/%y")


#2
#Building my default theme
my.default.theme <- theme_classic(base_size = 15) +
  theme(plot.background = element_rect(color='white',fill='white'),
        plot.title = element_text(color='black', size = 15),
        axis.text = element_text(color = 'black',size = 10), 
        panel.grid.minor = element_line(size = 0.5),
        panel.grid.major = element_line(size = 0.5),
        legend.background = element_rect(color='white', fill = 'white'),
        legend.position = "right",
        legend.title = element_text(color='black',size=10),
        legend.text = element_text(size = 10))

#Setting the theme as the default theme
theme_set(my.default.theme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: The depth of the lake has no correlation with the lake temparature recorded during July
Ha: The depth of the lake has correlation with the lake temparature recorded during July


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
#Wrangling the NTL-LTER dataset to follow the criteria
NTL_LTER_Sub  <- NTL_LTER %>%
  mutate(Month = month(sampledate))%>%
  filter (Month == "6") %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  drop_na()


#5
#Visualizing the relationship among the temperature and depth with a scatter plot
Temp.depth.scatt <- 
  ggplot(NTL_LTER_Sub, aes(x = depth, y = temperature_C)) +
  geom_point(alpha = 0.5) +
  geom_smooth (method="lm", color = "blue")+
  labs(x = "Depth (m)", y = "Temperature (°C)") +
  labs(title = "Scatterplot of Temperature by Depth")+
  ylim(0, 35)
  
print(Temp.depth.scatt) 

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: As the depth increases, the lake temperature decreases. The figure suggests temperature is negatively corrleated with depth (negative linearity). 


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
#Performing a linear regression on temperature and depth
LR.Temp.depth <- 
  lm(NTL_LTER_Sub$temperature_C ~ 
       NTL_LTER_Sub$depth)
summary (LR.Temp.depth)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:Since the adjusted R-squared is 0.6942, it suggests 69.42% of the variability in temperature is explained by changes in depth. The degrees of freedom on which this finding is based is 9501. The statistical significance of the result is < 2.2e-16, which is less than 0.05. This p-value suggests that there is significant correlation between temperature and depth. It is predicted 1.62 °C temperature will decrease for every 1m increase in depth.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
#Running an AIC to determine what set of explanatory variables is best suited to predict temperature
Temp.AIC <- lm(data = NTL_LTER_Sub, temperature_C ~ year4 + daynum + 
              depth)
step (Temp.AIC)
#Since the smaller the AIC value, the better. The model that includes all three explanatory variables (year4, daynum, depth) has the lowest value of AIC, which is 23932. Therefore, a multiple regression that includes year4, daynum, depth is best suited to predict temperature. 

#10
#Running a multiple regression on the recommended set of variables
Temp.Mul.Reg <- lm(data = NTL_LTER_Sub, temperature_C ~ year4 + daynum + depth)
summary(Temp.Mul.Reg)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression include year4, daynum, and depth. Since the adjusted R-squared is 0.7044, 70.44% of the observed variance this model can explain. Yes, this is an improvement over the model using only depth as the explanatory variable. The previous model using only depth can only explain 69.42% of the variability in temperature. However, the new multiple regression model can explain more variability in temperature by including additional explanatory variables, which are year4, daynum, and depth. 



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12


```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: 



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15


```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: 



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}

```

>Answer: 
